Trending assessment for nowcasts, measurements, and forecasts

### 2024-03-03 sections/03-trending.tex
Update on Overleaf.

As outlined in Section~\ref{sec:introduction}, we refer by trending to the consistency of an observed change with the change issued by another measurement, a nowcast, or a forecast.
In this section, we want to specify trending and lay out different methods of measuring and visualizing trending. 
In general, trending can be assessed for different lags $l$ separately, and the relevance of considered lags has to be determined based on the question at hand.
Throughout the section, we call a measurement, nowcast, or forecast a \textit{prediction} or \textit{signal}.
Before we introduce measures for assessing trending, we give remarks on computing the predicted change and notation in Section~\ref{subsec:notation}.
The remainder of the section is structured as follows.
Section~\ref{subsec:trending-basics} introduces trending basics and visualizes trending in a synthetical example.
Section~\ref{subsec:trending-measures} reviews conveyed quantification approaches from other fields.
The approaches are extended to the trending setting in Section~\ref{subsec:trending-noise}.
Section~\ref{subsec:trending-cond-prob} introduces a new graphical method of trending assessment and gives a brief introduction to bootstrap methods for computing confidence intervals for the measures.

@@

As noted above, trending refers to observed and predicted changes for a time lag $l$.
The observed change is straightforward to compute for measurements, nowcasts, and forecasts.
Let $\mathbf{y} = (y_t)_{t=0}^T$ denote the true values or gold standard measurements up to time $T$.
The sequence of observed change is then given by the differences of values in $\mathbf{y}$ with lag $l$, that is
\begin{equation}\label{eq:diffy}
    \diffylt = (y_{t+l} - y_t) \quad \text{for}\ t = 1, \dots, T - l.
\end{equation}
As noted in the introduction, the definition of predicted change depends on the context.
Conveyeing the computation for $y$ directly, the predicted change of signal would be the difference of the signal at time $t + l$ and the true value at $t$.
This approach lags in three practical aspects.
For nowcasting and forecasting, the true value at time $t$ might not be known with a considerable time lag.
In the example in Section~\ref{sec:application-covid}, for example, the time lag is more than 80 days.
Until the true values are made publicly available, the change cannot be computed and no actions can be taken based on it.
Thus, it is more practically relevant in this setting, to use updated signal values for time $t$ as the reference value.

Second, in the context of nowcasting or forecasting, several signals for the same time $t + l$ are often available with different times of generation.
Thus, it has to be specified in notation, which signal generated at which time is used.
We use the term \textit{issue time} for the generation time of a signal and \textit{target time} to denote the time the forecast applies to.

For measurement, usually, the use of a test method is evaluated against a gold standard to test whether the test method could replace the gold standard.
If the test method is assessed to be consistent with the gold standard, the gold standard measurement values are not available to compute the predicted change.
Thus, the predicted change is computed based on the test method's values at time $t$ and $t + l$.

Taking the three points together, the predicted change of a signal is based on the type of signal.
To account for several signals for a forecast time $t$ at different issue times $\tau$, we denote those by $x_{t | \tau}$.
For nowcasting, $t$ and $\tau$ are usually close and possibly negative, for forecasting, $t$ is usually larger than $\tau$.
For measurements, this distinction is not necessary.
Table~\ref{tab:notation} summarizes the notation for the different types of signals and the computation of the predicted change.

\begin{table}
    \centering
    \begin{tabularx}{0.75\textwidth}{l X}
        \toprule
        Application & Predicted change computation \\
        \midrule
        Measurement & $(x_{t + l} - x_t)_{t=0}^T \label{eq:diffxl_measure} $\\
        Nowcasting & $
\diffxl =
\begin{cases}
(x_{t+l|t+l} -x_{t|t+l})^T_{t=0} & \text{if } y_{t} \text{ is not known at time } t, \\
(x_{t+l|t+l} -y_{t})^T_{t=0} & \text{else}.
\end{cases} \label{eq:diffxl_nowcasting}
$ \\
        Forecasting & $
\diffxl =
\begin{cases}
(x_{t+l|t} - x_{t|t})^T_{t=0} & \text{if } y_{t} \text{ is not known at time } t, \\
(x_{t+l|t} - y_{t})^T_{t=0}  & \text{else}.
\end{cases} \label{eq:diffxl_forecasting}
$\\
        \bottomrule
    \end{tabularx}
    \caption{Computation of the predicted change in the different applications. The signals $x_{t | \tau}$ refer to values issued at time $\tau$ with a target time $t$. Note the different relations of issue and target time for nowcasts and forecasts. }
    \label{tab:notation}
\end{table}

@@

Throughout the remainder of the section, we use the notation of differences, $\diffxl$ and $\diffyl$.
While the computation of $\diffx$ differs in the different contexts, the trending assessment is similar.
In the following, we omit the difference's lag $l$ for the ease of notation; $\diffx$ and $\diffy$ refer to $\diffxl$ and $\diffyl$ for a common lag $l$.
As noted above, a signal is trending perfectly if all occurred directions of change are predicted correctly, that is, the sign of all elements of $\diffx$ and $\diffy$ coincide.
Perfect trending is hardly the case in practice, but a weaker requirement is needed.
Accordingly, when assessing the trending, we examine the statistical consistency of $\sign(\diffx)$ and $\sign(\diffy)$.
A simple yet insightful method is the four-quadrant plot.
For example, it is known in cardiac output measurement analysis~\parencite{Saugel2015,perrino1998intraoperative}. 
Thereby, the occurred changes are plotted together with the predicted changes, that is, $(\diffy_t, \diffx_t)$ for $t = 1, \dots, T$.
Thus, the x-axis of a four-quadrant plot shows the true value differences, whereas the y-axis displays prediction data differences.
Points in the green upper right and lower left quadrants reflect a correct trending for the respective time step, whereas points in the remaining red quadrants show incorrect predicted changes.
Figure~\ref{fig:trending_basic_4q} displays a basic four-quadrant plot.
Points 2, 3, 5, and 6 show trending, whereas points 1, 4, and 7 count for anti-trending behavior.
Figure~\ref{fig:trending_basic_4q_sample} shows a four-quadrant plot for a more extensive simulated data set with $T=1461$, for example, four years of daily data.
The data generation is described in Appendix~\ref{subsec:app-trending-data-generation} and will carry through this section.
The four-quadrant plot resembles a butterfly-like shape, with more data in the upper right and lower left quadrants than in the remaining quadrants.
The four-quadrant plot is intuitive to interpret, and the magnitude and direction of change are shown simultaneously.
It can be extended by including information on the date of difference in the point color.
In Figure~\ref{fig:trending_basic_4q_sample_color}, blue points refer to earlier measurements and turn green for larger $t$.
The lag is always one time step.
However, four-quadrant plots become crowded for larger datasets, and the sequential information on the differences is complex to assess thoroughly.
There are other visualization techniques, such as polar plots or the Bland-Altmann analysis.
However, they lack the clearness and intuition of the four-quadrant plot while adding no more information on the trending~\parencite{Saugel2015}.

@@

There are various adapted measures for unbalanced outcomes, for example, Cohen's $\kappa$ \parencite{Cohen1960} or those listed in \textcite[Table 3.3]{Jolliffe2012}.
Cohen's $\kappa$ is usually used to measure inter-rater agreement and takes into account the ratio of occurred agreement and the probability of agreement by chance.
Cohen's $\kappa$ reduces to rescaling the trending ratio in the case of a $2\times2$ table and balanced outcomes.
Balanced outcome hereby refers to approximately equal occurrences of values smaller and greater than zero for both $\diffx$ and $\diffy$.
Unbalanced outcomes of the differences are unlikely in our setting as they are obtained from differencing time series data.
Nevertheless, if the number of positive and negative $\diffy$ differs widely, the use of unbalanced-data-aware measures should be considered.

All measures can also be evaluated as a rolling estimate to detect changes in performance over time.
A rolling estimate is a sequence of estimates, each computed on a length-$w$-window of the data.
For the trending ratio a rolling estimate with a backwards-looking window is given by
\begin{equation*}
    \acc_t (\diffx, \diffy) \coloneqq \frac{\sum_{t^* = t-w + 1}^{t} \ind{\diffxt[t^*] \diffyt[t^*] > 0}}{w} \quad t = w-1, \dots, T.\label{eq:acc_rolling}
\end{equation*}
Figure~\ref{fig:trending_ratio_time_series} depicts a rolling window estimate of the trending ratio for the simulated data of Figures~\ref{fig:trending_basic_4q_sample} and~\ref{fig:trending_basic_4q_sample_color}.
Thus, the yearly course of the prediction's trending ratio can be detected.
The trending ability of the signal has a strong yearly, sinodal seasonality with higher trending ability peak after a quarter of a year and a low point after three quarters.
This seasonal behavior cannot be observed in the two four-quadrant plot as the high number of points overlays it or in the trending ratio.

@@

In applications, signal data is often not available for all time steps, for example, due to technical problems or delays in data transfer (see the examples in Sections~\ref{sec:application-covid} and~\ref{sec:application-forecasting}).
We refer to time steps, for which either signal or true values or both are not available, as missing values.
Data pairs with missing values can be excluded from the computation of the measures, so that the time indices with missing values are excluded from $\mathcal{T}$.
In general, the measures can be computed for data with missing values, if the missing values do not systematically occur.
Systematic missing values would lead to a biased estimate of the trending ratio, for example, if signal data is omitted in times of high change and thus uncertainty.
In this case, the measures should be computed for the time steps with available data and the results should be interpreted with caution.
Thus, inspection of missing data, for example, visually, should accompany applications with missing data.
In the applications in Sections~\ref{sec:application-covid} and~\ref{sec:application-forecasting}, only few missing values occur and we give information on the missing data by inspecting the missing data as lists.

@@

The measures are then estimators for probability of trending, given that the change is not in the exclusion area $e$, that is, $P(\diffxrv \diffyrv > 0 | \diffxrv \diffyrv \notin e)$.
The estimators can easily be adapted for various shapes of the exclusion area.
Figure~\ref{fig:trending_4q} visualizes different shapes of the exclusion area.
A rectangular shaped exclusion area, $e = \{(x, y) \in \R^2: (-\varepsilon_x \leq x \leq \varepsilon_x) \land (-\varepsilon_y \leq y \leq \varepsilon_y) \}$ for $\varepsilon_x, \varepsilon_y > 0$, leaves points out that are small in both components and are thus likely to be driven away from the zero point by noise.
Points, where at least true value or signal are unlikely to be zero, are not excluded.
In the example plot, only point 1 is excluded.
An exclusion area along one axis, for example, $e = \{(x, y) \in \R^2: (-\varepsilon_x \leq 0 \leq \varepsilon_x)\}$ for $\varepsilon_x > 0$, removes points in which one of the components could change the sign by a small amount of noise.
This particularly suits signals, where small amounts of noise are inevitable.
A cross-shaped exclusion area, $e = \{(x, y) \in \R^2: (-\varepsilon_x \leq x \leq \varepsilon_x) \lor (-\varepsilon_y \leq y \leq \varepsilon_y) \}$ for $\varepsilon_x, \varepsilon_y > 0$, along both axes accounts for likely sign reversal in both components.
For those two methods, points 1 and 7 or 1, 6, and 7 are excluded, respectively.

In addition to these straigt shapes, exclusion areas could have more difficult shapes, such as ellipsis.
Such a shape could, for example, be obtained by using a multivariate normal error model around the zero point and excluding points, that exceed a certain likelihood of being obtained by chance from the zero point.
These shapes are theoretically appealing, but the interpretation of the resulting estimators gets complex.
Thus, we use the simple exclusion areas above.

>>>>

As outlined in Section~\ref{sec:introduction}, we refer by trending to the consistency of an observed change with the change issued by another measurement, a nowcast, or a forecast.
In this section, we want to specify trending and lay out different methods of measuring and visualizing trending. 
In general, trending can be assessed for different lags $l$ separately, and the relevance of considered lags has to be determined based on the question at hand.
Throughout the section, we call a measurement, nowcast, or forecast a \textit{prediction} or \textit{signal}.
Before introducing measures for assessing trending, we give remarks on computing the predicted change and notation in Section~\ref{subsec:notation}.
The remainder of the section is structured as follows.
Section~\ref{subsec:trending-basics} introduces trending basics and visualizes trending in a synthetical example.
Section~\ref{subsec:trending-measures} reviews conveyed quantification approaches from other fields.
The approaches are extended to the trending setting in Section~\ref{subsec:trending-noise}.
Section~\ref{subsec:trending-cond-prob} introduces a new graphical method of trending assessment and briefly introduces bootstrap methods for computing confidence intervals for the measures.

@@

As noted above, trending refers to observed and predicted changes for a time lag $l$.
The observed change is straightforward to compute for measurements, nowcasts, and forecasts.
Let $\mathbf{y} = (y_t)_{t=0}^T$ denote the true values or gold standard measurements up to time $T$.
The sequence of observed change is then given by the differences of values in $\mathbf{y}$ with lag $l$, that is
\begin{equation}\label{eq:diffy}
    \diffylt = (y_{t+l} - y_t) \quad \text{for}\ t = 1, \dots, T - l.
\end{equation}
As noted in the introduction, the definition of predicted change depends on the context.
Conveying the computation for $y$ directly, the signal's predicted change would be the difference between the signal at time $t + l$ and the true value at $t$.
This approach lags in three practical aspects.
For nowcasting and forecasting, the true value at time $t$ might not be known with a considerable time lag.
In the example in Section~\ref{sec:application-covid}, for example, the time lag is more than 80 days.
Until the true values are publicly available, the change cannot be computed, and no actions can be taken based on it.
Thus, it is more practically relevant to use updated signal values for time $t$ as the reference value in this setting.

Second, in the context of nowcasting or forecasting, several signals for the same time $t + l$ are often available at different times of generation.
Thus, it has to be specified in notation which signal generated at which time is used.
We use the term \textit{issue time} for the generation time of a signal and \textit{target time} to denote the time the forecast applies to.

For measurement, usually, a test method is evaluated against a gold standard to test whether the test method could replace the gold standard.
If the test method is assessed to be consistent with the gold standard, the gold standard measurement values are not available to compute the predicted change.
Thus, the predicted change is computed based on the test method's values at time $t$ and $t + l$.

Taking the three points together, the signal's predicted change is based on the type of signal.
To account for several signals for a forecast time $t$ at different issue times $\tau$, we denote those by $x_{t | \tau}$.
For nowcasting, $t$ and $\tau$ are usually close and possibly negative; for forecasting, $t$ is usually larger than $\tau$.
For measurements, this distinction is not necessary.
Table~\ref{tab:notation} summarizes the notation for the different signal types and the predicted change computation.

\begin{table}
    \centering
    \begin{tabularx}{0.75\textwidth}{l X}
        \toprule
        Application & Predicted change computation \\
        \midrule
        Measurement & $(x_{t + l} - x_t)_{t=0}^T \label{eq:diffxl_measure} $\\
        Nowcasting & $
\diffxl =
\begin{cases}
(x_{t+l|t+l} -x_{t|t+l})^T_{t=0} & \text{if } y_{t} \text{ is not known at time } t, \\
(x_{t+l|t+l} -y_{t})^T_{t=0} & \text{else}.
\end{cases} \label{eq:diffxl_nowcasting}
$ \\
        Forecasting & $
\diffxl =
\begin{cases}
(x_{t+l|t} - x_{t|t})^T_{t=0} & \text{if } y_{t} \text{ is not known at time } t, \\
(x_{t+l|t} - y_{t})^T_{t=0}  & \text{else}.
\end{cases} \label{eq:diffxl_forecasting}
$\\
        \bottomrule
    \end{tabularx}
    \caption{Computation of the predicted change in the different applications. The signals $x_{t | \tau}$ refer to values issued at $\tau$ with a target time $t$. Note the different relations of issue and target time for nowcasts and forecasts. }
    \label{tab:notation}
\end{table}

@@

Throughout the remainder of the section, we use the notation of differences, $\diffxl$ and $\diffyl$.
While the computation of $\diffx$ differs in different contexts, the trending assessment is similar.
In the following, we omit the difference's lag $l$ for the ease of notation; $\diffx$ and $\diffy$ refer to $\diffxl$ and $\diffyl$ for a common lag $l$.
As noted above, a signal is trending perfectly if all occurred directions of change are predicted correctly; that is, the sign of all elements of $\diffx$ and $\diffy$ coincide.
Perfect trending is hardly the case in practice, but a weaker requirement is needed.
Accordingly, when assessing the trending, we examine the statistical consistency of $\sign(\diffx)$ and $\sign(\diffy)$.
A simple yet insightful method is the four-quadrant plot.
For example, it is known in cardiac output measurement analysis~\parencite{Saugel2015,perrino1998intraoperative}. 
Thereby, the occurred changes are plotted together with the predicted changes, that is, $(\diffy_t, \diffx_t)$ for $t = 1, \dots, T$.
Thus, the x-axis of a four-quadrant plot shows the true value differences, whereas the y-axis displays prediction data differences.
Points in the green upper right and lower left quadrants reflect a correct trending for the respective time step, whereas points in the remaining red quadrants show incorrect predicted changes.
Figure~\ref{fig:trending_basic_4q} displays a basic four-quadrant plot.
Points 2, 3, 5, and 6 show trending, whereas points 1, 4, and 7 count for anti-trending behavior.
Figure~\ref{fig:trending_basic_4q_sample} shows a four-quadrant plot for a more extensive simulated data set with $T=1461$, for example, four years of daily data.
The data generation is described in Appendix~\ref{subsec:app-trending-data-generation} and will carry through this section.
The four-quadrant plot resembles a butterfly-like shape, with more data in the upper right and lower left quadrants than in the remaining quadrants.
The four-quadrant plot is intuitive to interpret, and the magnitude and direction of change are shown simultaneously.
It can be extended by including information on the date of difference in the point color.
In Figure~\ref{fig:trending_basic_4q_sample_color}, blue points refer to earlier measurements and turn green for larger $t$.
The lag is always one time step.
However, four-quadrant plots become crowded for larger datasets, and the sequential information on the differences is complex to assess thoroughly.
There are other visualization techniques, such as polar plots or the Bland-Altmann analysis.
However, they lack the clearness and intuition of the four-quadrant plot while adding no more information on the trending~\parencite{Saugel2015}.

@@

There are various adapted measures for unbalanced outcomes, for example, Cohen's $\kappa$ \parencite{Cohen1960} or those listed in \textcite[Table 3.3]{Jolliffe2012}.
Cohen's $\kappa$ is usually used to measure inter-rater agreement and takes into account the ratio of occurred agreement and the probability of agreement by chance.
Cohen's $\kappa$ reduces to rescaling the trending ratio in the case of a $2\times2$ table and balanced outcomes.
Balanced outcome refers to approximately equal occurrences of values smaller and greater than zero for both $\diffx$ and $\diffy$.
Unbalanced outcomes of the differences are unlikely in our setting as they are obtained from differencing time series data.
Nevertheless, if the number of positive and negative $\diffy$ differs widely, the use of unbalanced-data-aware measures should be considered.

All measures can also be evaluated as a rolling estimate to detect changes in performance over time.
A rolling estimate is a sequence of estimates, each computed on a length-$w$-window of the data.
For the trending ratio, a rolling estimate with a backward-looking window is given by
\begin{equation*}
    \acc_t (\diffx, \diffy) \coloneqq \frac{\sum_{t^* = t-w + 1}^{t} \ind{\diffxt[t^*] \diffyt[t^*] > 0}}{w} \quad t = w-1, \dots, T.\label{eq:acc_rolling}
\end{equation*}
Figure~\ref{fig:trending_ratio_time_series} depicts a rolling window estimate of the trending ratio for the simulated data of Figures~\ref{fig:trending_basic_4q_sample} and~\ref{fig:trending_basic_4q_sample_color}.
Thus, the yearly course of the prediction's trending ratio can be detected.
The trending ability of the signal has a strong yearly, sinus-shaped seasonality with higher trending ability peak after a quarter of a year and a low point after three quarters.
This seasonal behavior cannot be observed in the two four-quadrant plots as the high number of points overlays it or in the trending ratio.

@@

In applications, signal data is often not available for all time steps, for example, due to technical problems or delays in data transfer (see the examples in Sections~\ref{sec:application-covid} and~\ref{sec:application-forecasting}).
We refer to time steps, for which either signal or true values or both are unavailable, as missing values.
Data pairs with missing values can be excluded from the computation of the measures so that the time indices with missing values are excluded from $\mathcal{T}$.
In general, the measures can be computed for data with missing values if the missing values do not systematically occur.
Systematic missing values would lead to a biased estimate of the trending ratio, for example, if signal data is omitted in times of high change and thus uncertainty.
In this case, the measures should be computed for the time steps with available data, and the results should be interpreted cautiously.
Thus, inspection of missing data should accompany applications with missing data, for example, through a visual assessment.
In the applications in Sections~\ref{sec:application-covid} and~\ref{sec:application-forecasting}, only a few missing values occur, and we give information on the missing data by inspecting the missing data as lists.

@@

The measures are then estimators for the probability of trending, given that the change is not in the exclusion area $e$, that is, $P(\diffxrv \diffyrv > 0 | \diffxrv \diffyrv \notin e)$.
The estimators can easily be adapted for various shapes of the exclusion area.
Figure~\ref{fig:trending_4q} visualizes different shapes of the exclusion area.
A rectangularly shaped exclusion area, $e = \{(x, y) \in \R^2: (-\varepsilon_x \leq x \leq \varepsilon_x) \land (-\varepsilon_y \leq y \leq \varepsilon_y) \}$ for $\varepsilon_x, \varepsilon_y > 0$, leaves points out that are small in both components and are thus likely to be driven away from the zero point by noise.
Points where at least the true value or signal is unlikely to be zero are not excluded.
In the example plot, only point 1 is excluded.
An exclusion area along one axis, for example, $e = \{(x, y) \in \R^2: (-\varepsilon_x \leq 0 \leq \varepsilon_x)\}$ for $\varepsilon_x > 0$, removes points in which one of the components could change the sign by a small amount of noise.
This particularly suits signals where small amounts of noise are inevitable.
A cross-shaped exclusion area, $e = \{(x, y) \in \R^2: (-\varepsilon_x \leq x \leq \varepsilon_x) \lor (-\varepsilon_y \leq y \leq \varepsilon_y) \}$ for $\varepsilon_x, \varepsilon_y > 0$, along both axes accounts for likely sign reversal in both components.
For those two methods, points 1 and 7 or 1, 6, and 7 are excluded, respectively.

In addition to these straight shapes, exclusion areas such as ellipsis could have more difficult shapes.
Such a shape could, for example, be obtained by using a multivariate normal error model around the zero point and excluding points, that exceed a certain likelihood of being obtained by chance from the zero point.
These shapes are theoretically appealing, but the interpretation of the resulting estimators gets complex.
Thus, we use the simple exclusion areas above.
